---
title: "An Introduction to ggplot2"
author: "Daina Bouquin"
output: html_document
---

All code associated with this presentation is available on [GitHub][1].

************

*Objectives:*  
1. Familiarize you with the `ggplot2` R package  
2. Introduce you to the `qplot()` function  
3. Briefly demonstrate the more advanced `ggplot()` function  

************
<h3> What is `ggplot2`?</h3>
  
Created by Hadley Wickham, `ggplot2` is an incredibly popular and powerful graphics language for creating elegant and complex plots. Origianlly based on Leland Wilkinson's *The Grammar of Graphics*, `ggplot2` allows you to create graphs that represent both univariate and multivariate numerical and categorical data in a straightforward manner. Grouping can be represented by color, symbol, size, and transparency.

Data visualizations that would require dozens of lines of code on other platforms or with base R graphics, only require a few quick lines once we have installed `ggplot2`. For example:
  
```{r}     
library(gcookbook)    # R Graphics Cookbook - package contains the data for this first example
library(grid)         # Need for arrow() function
library(ggplot2)
```
```{r, echo = FALSE} 
# Data pre-processing steps

islice <- subset(isabel, z == min(z)) 

# keep 1 out of every 'by' values in vector x
every_n <- function(x, by = 2) {
  x <- sort(x)
  x[seq(1, length(x), by = by)]
}

# keep 1 out of every 4 values in x and y
keepx <- every_n(unique(isabel$x), by = 4)
keepy <- every_n(unique(isabel$y), by = 4)

# keep only those rows where x value is in keepx and y value is in keepy
islicesub <- subset(islice, x %in% keepx & y %in% keepy)

# calculate speed components
islicesub$speedxy <- sqrt(islicesub$vx^2 + islicesub$vy^2)

# Get USA map data
usa <- map_data("usa")
```

```{r, warning=FALSE}
ggplot(islicesub, aes(x=x, y=y)) +
  geom_segment(aes(xend = x+vx/50, yend = y+vy/50, colour = speed),
               arrow = arrow(length = unit(0.1,"cm")), size = 0.6) +
  scale_colour_continuous(low="grey80", high="darkred") +
  geom_path(aes(x=long, y=lat, group=group), data=usa) +
  coord_cartesian(xlim = range(islicesub$x), ylim = range(islicesub$y))
```
  
************  
  
    
<h3>`qplot()`: the "Quick Plot"</h3>

**All qplots contain:** 
  
* aesthetics: size, shape, color, etc.
* geoms: points, lines, bars, etc.
  
Plots are draw on a specific coordinate system   
   
Factors are important for plotting subsets  
  
Data should be organized in a dataframe      
    
We will use the dataset `mpg` for our example. `mpg` is part of the ggplot2 package.
```{r}
str(mpg)
````
str(mpg)

**We will first examine the "base" plotting functions of R for comparison**
  
With base functions we start with a blank canvas and build up piece by piece. This can get difficult pretty quickly if we're trying to make anything complicated.

Note that the `$` allows us to point to the column names in the dataset
```{r}
plot(mpg$displ, mpg$hwy)
```

Now try the `qplot()` from `ggplot2` 
```{r}
qplot(displ, hwy, data = mpg)
```

What do you notice that's different?  
  
We'll start adding aesthetics now. Here we use the `color` parameter to indicate that different levels of drive variable (`drv`) should each be shown in a different color. This is possible through the use of factors

```{r}
qplot(displ, hwy, data = mpg, color = drv)
```

*Note how the legend gets placed automatically*- it can be difficult to customize `qplot()` so it's important to learn the higher functionality of ggplot2 as well.  
   
Lets try some more options for `qplot()`

Lets add a geom (statistic) summary of data to the plot to smooth the data and see general trends. The grey zone indicates my .95 confidence interval.


```{r}
qplot(displ, hwy, data = mpg, geom = c("point", "smooth"), method = lm)
```

We can use `qplot()` to make all kinds of plots including histograms
Here's an example of a histogram using the "fill" argument. Note that we do not need to specify the `"histogram"` parameter but I have done so here for clarity. Entering just one variable (`hwy`) will create a histogram. 
```{r}
qplot(hwy, data = mpg, geom= "histogram", fill = drv, binwidth = 1)
```

As an alternative to approaches like color coding, ggplot2 let's us use facets to split variables by their levels.  

The variable in the left facets argment specifies rows, and right arguments specifies columns (drv in this case). The data arguments are separated by  tilda 
```{r} 
qplot(displ, hwy, data = mpg, facets = . ~ drv)
```

How about we try a histogram?
```{r}
qplot(hwy, data = mpg, geom = "histogram", facets = drv ~ . , binwidth = 1)
```

It can be difficult to customize `qplot()` though remember, so learning full `ggplot2` functionality is important; for this reason we'll briefly look at `ggplot()`.
  
************  
  
<h3>`ggplot()` in Brief </h3>  

`ggplot()` is a bit more advanced than qplot, but it's incredibly robust and customizable.   
  
You create a `ggplot()` object, and you can then add "layers" to your plot to customize it.

```{r}

mpg_1 <- ggplot(mpg, aes(drv, hwy))
```
  
Now if I try running this I'll get an error because there are no "layers" in the plot  
```{r, error=TRUE}
mpg_1
```
  
Note though that I have created a ggplot() object, which means I can run functions on it.   
For example, I can take a summary of the object that tells me what's going on with it:

```{r}
summary(mpg_1)
```
   
So what I need to do is add some "layers"   
Here I've added the `geom_violin()` function. Violin plots are ordinarily quite difficult to create but here it's easy. 

```{r}
mpg_vplot <- mpg_1 + geom_violin()
mpg_vplot
```
  
If I want to show more granularly the distribution of the points, I can just literally "add" more geoms.   
   
*Note I'm adjusting my spacing here to make the code easier to read.*
   
```{r}
mpg_vplot <- mpg_1 + geom_violin() + 
  geom_point() 

mpg_vplot
```
  
We can also start to play with color packages and different types of geoms and parameters from here.
  
For example, we can define a color blind friendly palette using hexadecimal codes or "hex color codes" and create a faceted histogram to show the count of cars by `drv` and `cyl`:
  
```{r}
# color blind friendly palette:
cbPalette <- c("#E69F00", "#56B4E9", "#009E73")

scale_fill_manual(values=cbPalette)

mpg_2 <- ggplot(mpg, aes(drv, fill = drv)) 

cb_plot <- mpg_2  + 
  geom_bar(width = .5) + 
  facet_wrap(~ cyl) + 
  scale_fill_manual(values=cbPalette)

cb_plot

```

[1]: https://github.com/dbouquin

